---
title: "Analysing Severe Adverse Effects"
author: "Chhiring Lama"
date: "2026-03-23"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{lipsum}
  - \pagestyle{fancy}
  - \fancyhead[R]{\thepage}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \usepackage{enumitem}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gtsummary)
library(performance)
library(readxl)
library(naniar)
library(gt)
library(kableExtra)
library(lme4)
library(nlme)
library(ggpubr)
library(sjPlot)
library(DHARMa)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

options(scipen = 999)
```

### data wrangling
```{r}
sae_df <- read_xlsx("data/Q2b.xlsx") |> 
  janitor::clean_names() |> 
  mutate(sae = ifelse(sae == 1, "Yes", "No") |> as.factor(), 
         group = ifelse(group == 1, "Vaccine", "Control") |> as.factor())
#time = as.factor(time), 
         

bl_df <- read_xlsx("data/Q2b_BL.xlsx") |> 
  janitor::clean_names() |> 
  mutate(sex = ifelse(sex == 1, "Male", "Female"), 
         sex = fct_relevel(sex, "Male"))

combined_dataset <- sae_df |> 
  full_join(bl_df, by = "id") 

combined_dataset_wide <- combined_dataset |> 
  pivot_wider(names_from = "time", 
              values_from = "sae", 
              names_prefix = "month_")
```

Summary Table
```{r}
table1  <- combined_dataset_wide %>%
  dplyr::select(group:age) %>%
  tbl_summary(
    by = group,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",    # drop min/max to save space
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      #time ~ 'Time (months)',
      age ~ "Age (years)", 
      sex ~ "Sex",
      #sae ~ "Severe Adverse Events (Yes)", 
      site ~ "Number of Sites")
  ) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels() %>%
  modify_caption("**Baseline Characteristics by Treatment Group**")
# table1v2 <- table1 %>%
#   as_gt() %>%
#   gt::tab_options(
#     data_row.padding = px(2),
#     table.width      = pct(100),
#     table.border.top.color = "black",
#     table.border.bottom.color = "black",
#     heading.border.bottom.color = "black",
#     table.font.color = "black",
#     column_labels.border.bottom.style = "none", # Remove header line
#     table_body.border.bottom.style = "none",    # Remove body line
#     table_body.border.top.style = "none",
#     table.border.top.style = "none",
#     table.font.names = "arial",
#     table.font.size = "12px" 
#   ) |> 
#   cols_width(
#     label      ~ px(150),
#     everything() ~ px(80)
#   ) |> 
#   gt::tab_style(
#     style = gt::cell_borders(), 
#     locations = gt::cells_body(columns = everything()))  |> 
#   gt::gtsave(filename = "figures_tables/summarytab.pdf")
#system("pdfcrop figures_tables/summarytab.pdf figures_tables/summarytab.pdf")

table1 |> 
  as_gt() |> 
  cols_width(
    label      ~ px(150),
    everything() ~ px(80)
  ) |>
  gt::gtsave(filename = "figures_tables/summarytab.pdf")

table1 |> 
  as_kable()
```


```{r}
pdf("figures_tables/age_range.pdf", width = 6, height = 3)
ggplot(aes(x = age, fill = group), data = combined_dataset_wide) +
  geom_density(alpha = 0.4) +
  labs(x = "Age", y = "Density", fill = "Tx group") +
  theme_classic()
dev.off()
```

Checking for missing data pattern (if any)
```{r, out.width="100%", warning=FALSE}
fig2a <- vis_miss(combined_dataset_wide) + 
  theme(axis.text.x = element_text(vjust = 0))
ggsave("figures_tables/missingdata_info.png", fig2a, height = 5, width =8)
```

Missingness only seen in outcome data (SAE) and more so for 2nd and third timepoints

Missingness by gender
```{r, fig.height = 3, fig.width=6, warning=FALSE}
tx_gender_missing <- combined_dataset |> 
  group_by(sex, time) |> 
  mutate(total = length(unique(id))) |> 
  ungroup() |> 
  drop_na() |> 
  group_by(sex, time) |> 
  summarize (missing_pct = (total[1] - length(unique(id)))*100/total[1])

missing_by_gender <- ggplot(aes(x = time, y = missing_pct, group = sex, color = sex), 
       data = tx_gender_missing) +
  geom_point() +
  scale_x_continuous(limits = c(1, 3))+
  geom_line() +
  labs(x = "Time after Vaccine (in months)", y = "Missing Data %", color = "Sex") +
  theme_classic()

ggsave("figures_tables/missingness_by_gender.pdf", missing_by_gender, height = 3, width =5)
#system("pdfcrop figures_tables/missingness_by_gender.pdf figures_tables/missingness_by_gender.pdf")
missing_by_gender
```

### Characterise missingness pattern
Number of drop outs 
```{r, warning=FALSE}
missingpattern <- combined_dataset_wide|> 
  mutate(pattern = case_when(is.na(month_3) == FALSE & is.na(month_2) == FALSE & 
                               is.na(month_1) == FALSE ~ "Complete", ##all or at least 2 where one must be from month 3
                             is.na(month_3) == FALSE & is.na(month_2) == TRUE & 
                               is.na(month_1) == FALSE ~ "Complete",
                             is.na(month_3) == FALSE & is.na(month_2) == FALSE & 
                               is.na(month_1) == TRUE ~ "Complete",
                             is.na(month_3) == TRUE & is.na(month_2) == TRUE &
                               is.na(month_1) == TRUE ~ "Assessment Dropout", 
                             is.na(month_3) == TRUE & is.na(month_2) == TRUE &
                               is.na(month_1) == FALSE ~ "Early Dropout", 
                             is.na(month_3) == TRUE & is.na(month_2) == FALSE &
                               is.na(month_1) == TRUE ~ "Only 2nd Visit", 
                             is.na(month_3) == FALSE & is.na(month_2) == TRUE &
                               is.na(month_1) == TRUE ~ "Only 3rd Visit", 
                             is.na(month_3) == TRUE ~ "Late Dropout"),  
         pattern = fct_relevel(pattern, "Assessment Dropout", "Early Dropout", 
                               "Late Dropout"))
missingpattern1 <- missingpattern |> 
  dplyr::select(c(sex, pattern)) |> 
  tbl_summary(
    by = pattern,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",    # drop min/max to save space
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      sex ~ "Sex")
  ) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels()

missingpattern1 |> 
  as_gt() |> 
  cols_width(
    label      ~ px(76),
    everything() ~ px(64)
  ) |>
  tab_options(
    table.font.size = px(10),
    heading.title.font.size = px(11),
    column_labels.font.size = px(10),
    table.font.names = "Arial"
  ) |> 
  gt::gtsave(filename = "figures_tables/missingpattern1.pdf")

missingpattern1 |> 
  as_kable()
```

```{r, warning=FALSE}
missingpatternv2 <- missingpattern |> 
  dplyr::select(c(group, pattern)) |> 
  tbl_summary(
    by = group,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",    # drop min/max to save space
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      pattern ~ "Missingness Pattern")
  ) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels()

missingpatternv2 |> 
  as_gt() |> 
  gt::gtsave(filename = "figures_tables/missingpatternv2.pdf")

missingpatternv2 |> 
  as_kable()
```
Only 3.7% missingness, ignoring for now

EDA for SAE
```{r, fig.height = 4, fig.width=9, warning=FALSE}
sae_pct <- combined_dataset |> 
  drop_na() |> 
  group_by(group, time, site, sex) |> 
  mutate(total = n()) |> 
  ungroup() |> 
  dplyr::filter(sae == "Yes") |>  
  group_by(group, time, site, sex) |> 
  summarise (sae_pct = n()*100/total[1]) |> 
  mutate(group_time = paste(time, group, sep = "_")) 

sae_pct_by_gender_persite <- ggplot(aes(x = group_time, y = sae_pct, color = group), 
       data = sae_pct) +
  geom_boxplot() +
  geom_jitter(alpha = 0.4) +
  stat_compare_means(comparisons = list(c("1_Control", "1_Vaccine"), 
                                        c("2_Control", "2_Vaccine"), 
                                        c("3_Control", "3_Vaccine")), 
                     label = "p.adj.format",
                     p.adjust.method = "bonferroni") +
  labs(x = "Time after Vaccine (in months) + Tx Group", y = "Severe Adverse Event% (Per Site)", color = "Group") +
  facet_wrap(~sex) +
  theme_classic()

ggsave("figures_tables/sae_pct_by_gender_persite.pdf", sae_pct_by_gender_persite, height = 5, width =11)
#system("pdfcrop figures_tables/sae_pct_by_gender_persite.pdf figures_tables/sae_pct_by_gender_persite.pdf")
sae_pct_by_gender_persite
```

```{r, fig.height = 3, fig.width=5, warning=FALSE}
sae_pct <- combined_dataset |> 
  drop_na() |> 
  group_by(group, time) |> 
  mutate(total = n()) |> 
  ungroup() |> 
  dplyr::filter(sae == "Yes") |>  
  group_by(group, time) |> 
  mutate (sae_pct = n()*100/total[1])

sae_pct_by_gender <- ggplot(aes(x = time, y = sae_pct, group = group, color = group), 
       data = sae_pct) +
  geom_point() +
  scale_x_continuous(limits = c(1, 3), breaks = c(1, 2, 3))+
  geom_line() +
  labs(x = "Time after Vaccine (in months)", y = "Severe Adverse Event %", color = "Group") +
  theme_classic()

ggsave("figures_tables/sae_pct_by_gender.pdf", sae_pct_by_gender, height = 3, width =5)
#system("pdfcrop figures_tables/sae_pct_by_gender.pdf figures_tables/sae_pct_by_gender.pdf")
sae_pct_by_gender
```

```{r, fig.height = 3, fig.width=7, warning=FALSE}
sae_pct_by_group <- combined_dataset |> 
  drop_na() |> 
  group_by(sex, time, group) |> 
  mutate(total = n()) |> 
  ungroup() |> 
  dplyr::filter(sae == "Yes") |>  
  group_by(sex, time, group) |> 
  mutate (sae_pct = n()*100/total[1])

sae_pct_by_gender_group <- ggplot(aes(x = time, y = sae_pct, group = group, color = group), 
       data = sae_pct_by_group) +
  geom_point() +
  scale_x_continuous(limits = c(1, 3), breaks = c(1, 2, 3))+
  geom_line() +
  facet_wrap(~sex) +
  labs(x = "Time after Vaccine (in months)", y = "Severe Adverse Event %", color = "Group") +
  theme_classic()

ggsave("figures_tables/sae_pct_by_gender_group.pdf", sae_pct_by_gender_group, height = 3, width =7)
#system("pdfcrop figures_tables/sae_pct_by_gender_group.pdf figures_tables/sae_pct_by_gender_group.pdf")
sae_pct_by_gender_group
```

###sae by age per group
```{r, fig.height = 3, fig.width=7, warning=FALSE}
sae_by_age <- ggplot(aes(x = sae, y = age, color = sae), 
       data = combined_dataset |> drop_na(sae) |> 
  mutate(group_time = paste(group, time, sep = "_"))) +
  geom_boxplot() +
  labs(x = "SAE", y = "Age", color = "Group") +
  facet_wrap(~group_time, ncol = 3)+
  scale_y_continuous(limits = c(10, 120)) +
  stat_compare_means() +
  theme_classic()

ggsave("figures_tables/sae_by_age.pdf", sae_by_age, height = 6, width =8)
#system("pdfcrop figures_tables/sae_pct_by_gender_group.pdf figures_tables/sae_pct_by_gender_group.pdf")
sae_by_age
```


Interaction term not needed
read on correlation structure, addin two separate random intercept etc. 

PMM: run GLMM per pattern
```{r}
long_df_w_pattern <- missingpattern |> 
  pivot_longer(cols = month_1:month_3, 
               names_to = "time", 
               values_to = "sae") |> 
  mutate(time = as.character(time))

pattern_list <- setdiff(na.omit(as.character(unique(long_df_w_pattern$pattern))), 
                        "Assessment Dropout")
run_glmm <- lapply(pattern_list, function(pat){
  df <- long_df_w_pattern |> dplyr::filter(pattern == pat) |> 
    mutate(sae = as.character(sae), 
           time = as.character(time)) |> 
    drop_na()
  if (length(unique(na.omit(df$sae))) > 1 & length(unique(df$time)) > 1){
    print(pat)
    print(unique(df$time))
    df <- df |> 
      mutate(sae = as.factor(sae), 
             time = as.factor(time))
    fit_PMM <- glmer(sae ~ group + time + sex + age + (1 | site),
                        data = df, family = binomial, nAGQ = 0)
  }
  if (length(unique(na.omit(df$sae))) > 1 & length(unique(df$time)) == 1){
    print(pat)
    print(unique(df$time))
    df <- df |> 
      mutate(sae = as.factor(sae))
    fit_PMM <- glmer(sae ~ group + sex + age + (1 | site),
                        data = df, family = binomial, nAGQ = 0)
  }
  if (length(unique(na.omit(df$sae))) == 1 & length(unique(df$time)) == 1){
    fit_PMM <- NULL
  }
  return(fit_PMM)
})

names(run_glmm) <- pattern_list
run_glmm <- Filter(Negate(is.null), run_glmm)
```

##Compare coefficients
```{r}
pairs <- combn(names(run_glmm), 2, simplify = FALSE)

lapply(pairs, function(compare){
  first_fit <- run_glmm[[compare[1]]]
  second_fit <- run_glmm[[compare[2]]]
  coef_first <- fixef(first_fit)
  coef_second <- fixef(second_fit)
  
  vcov_first <- vcov(first_fit)
  vcov_second <- vcov(second_fit)
  
  overlapping_terms <- intersect(names(coef_first), 
                                 names(coef_second))
  res_complete_late <- lapply(overlapping_terms, function(term){
  
    beta1 <- coef_first[term]
    beta2 <- coef_second[term]
  
    var1 <- vcov_first[term, term]
    var2 <- vcov_second[term, term]
  
    z <- (beta1 - beta2) / sqrt(var1 + var2)
    p <- 2 * (1 - pnorm(abs(z)))
    
    data.frame(
      `Group` = paste0(compare[1], " vs ", compare[2]), 
      `Beta1` = beta1, `Beta2` = beta2, `z` = z, `P-value` = p
    )
  })
  comp <- knitr::kable(as.data.frame(do.call(rbind, res_complete_late)), digits = 3) |> 
    kable_styling(latex_options = c("striped", "hold_position"), font_size = 12)
  print(comp)
  #save_kable(comp, file = paste0("figures_tables/coef_comparison", compare[1], compare[2], ".pdf"))
})
```

Modeling: GLMM to model for odds of SAE at each timepoint
```{r, warning=FALSE}
combined_dataset2 <- combined_dataset |> 
  mutate(time = as.factor(time))
glmm <- glmer(sae ~ group*time + sex + age + (1 | site) + (1 | id) , 
               family = binomial, data = combined_dataset2, nAGQ = 0)

```
Singularity issue when including site +id or site(id)


Final Model:
$$
\begin{aligned}
\widehat{logit}(P(SAE_{ijk} = 1)) & = \beta_0 + \beta_1.I(Group_{ijk} \text{ = Vaccine)} + \beta_2.I(Month_{ijk} \text{= 2)}  + \\
&\quad \beta_3.I(Month_{ijk} \text{ = 3)} + \beta_4.I(Sex_{ijk} \text{ = Female)} + \beta_5.Age_{ijk} + b_{i} + c_{j}, \\
&\quad i = (1, 2, \ldots, 100),j = (1, 2, \ldots, 41250), \ k = 1, 2, 3, \\
&\quad b_{i} \sim N(0, {{\sigma}^2}_b)
\end{aligned}
$$

Result form GLMM2 (Main Effects)
```{r}
glmm_ad_table <- tbl_regression(
  glmm,
  exponentiate = TRUE, 
  include = c("(Intercept)", "group", "time","sex", "age"),
  intercept = TRUE,
  label = list(
    group ~ "Treatment",
    time ~ "Time (month)",
    sex ~ "Sex", 
    age ~ "Age"),
  estimate_fun = function(x) {
  ifelse(abs(x) < 1e-2,
         formatC(x, format = "e", digits = 2),
         style_number(x, digits = 2))
}) %>%
  bold_p() %>%
  bold_labels() 
glmm_ad_table
```

ANOVA for interaction term
```{r}
glmm_ad_table <- tbl_regression(
  glmm,
  exponentiate = TRUE, 
  include = c("group:time"),
  intercept = TRUE,
  label = list(
    `group:time` ~ "Treatment:Time (month)"),
  estimate_fun = function(x) {
  ifelse(abs(x) < 1e-2,
         formatC(x, format = "e", digits = 2),
         style_number(x, digits = 2))
}) %>%
  bold_p() %>%
  bold_labels() |> 
  add_global_p() ##run anova for interaction term
glmm_ad_table
```



# ```{R}
# mem.maxVSize(30*1024^3)
# plot_model(glmm2, type = "pred", terms = c("age [25:75]", "time", "group"), ci.lvl = 0.95) +
#   labs(title = "",  
#        y = "Predicted Probability",
#      x = "Age", 
#      color = "Time") +
#   scale_y_continuous(labels = function(x) formatC(x, digits = 3, format = "g")) +
#   theme_classic()
# ggsave("figures_tables/predicted_prob.pdf", height = 4, width =8)
# ```

##Diagnostics
```{r}
sim_ad <- simulateResiduals(fittedModel = glmm)
pdf("figures_tables/model_diganostics.pdf", width = 7, height = 5)
plot(sim_ad)
dev.off()
plot(sim_ad)
```

```{r, fig.height=12, fig.width=10,  fig.align='left'}
#check_model(glmm)
# pdf("figures_tables/model_diganostics2.pdf", width = 10, height = 12)
# check_model(glmm)
# dev.off()
```

