---
title: "Analysing Severe Adverse Effects"
author: "Chhiring Lama"
date: "2026-03-23"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{lipsum}
  - \pagestyle{fancy}
  - \fancyhead[R]{\thepage}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \usepackage{enumitem}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gtsummary)
library(performance)
library(readxl)
library(naniar)
library(gt)
library(kableExtra)
library(lme4)
library(nlme)
library(ggpubr)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

options(scipen = 999)
```

### data wrangling
```{r}
sae_df <- read_xlsx("data/Q2b.xlsx") |> 
  janitor::clean_names() |> 
  mutate(sae = ifelse(sae == 1, "Yes", "No") |> as.factor(), 
         group = ifelse(group == 1, "Vaccine", "Control") |> as.factor())
#time = as.factor(time), 
         

bl_df <- read_xlsx("data/Q2b_BL.xlsx") |> 
  janitor::clean_names() |> 
  mutate(sex = ifelse(sex == 1, "Male", "Female"), 
         sex = fct_relevel(sex, "Male"))

combined_dataset <- sae_df |> 
  full_join(bl_df, by = "id") 

combined_dataset_wide <- combined_dataset |> 
  pivot_wider(names_from = "time", 
              values_from = "sae", 
              names_prefix = "month_")
```

Summary Table
```{r}
table1  <- combined_dataset %>%
  drop_na() |> 
  dplyr::select(time:age) %>%
  tbl_summary(
    by = group,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",    # drop min/max to save space
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      time ~ 'Time (months)',
      age ~ "Age (years)", 
      sex ~ "Sex",
      sae ~ "Severe Adverse Events (Yes)", 
      site ~ "Number of Sites")
  ) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels() %>%
  modify_caption("**Baseline Characteristics by Treatment Group**")
# table1v2 <- table1 %>%
#   as_gt() %>%
#   gt::tab_options(
#     data_row.padding = px(2),
#     table.width      = pct(100),
#     table.border.top.color = "black",
#     table.border.bottom.color = "black",
#     heading.border.bottom.color = "black",
#     table.font.color = "black",
#     column_labels.border.bottom.style = "none", # Remove header line
#     table_body.border.bottom.style = "none",    # Remove body line
#     table_body.border.top.style = "none",
#     table.border.top.style = "none",
#     table.font.names = "arial",
#     table.font.size = "12px" 
#   ) |> 
#   cols_width(
#     label      ~ px(150),
#     everything() ~ px(80)
#   ) |> 
#   gt::tab_style(
#     style = gt::cell_borders(), 
#     locations = gt::cells_body(columns = everything()))  |> 
#   gt::gtsave(filename = "figures_tables/summarytab.pdf")
#system("pdfcrop figures_tables/summarytab.pdf figures_tables/summarytab.pdf")

table1 |> 
  as_gt() |> 
  cols_width(
    label      ~ px(150),
    everything() ~ px(80)
  ) |>
  gt::gtsave(filename = "figures_tables/summarytab.pdf")

table1 |> 
  as_kable()
```


Checking for missing data pattern (if any)
```{r, out.width="100%", warning=FALSE}
fig2a <- vis_miss(combined_dataset_wide) + 
  theme(axis.text.x = element_text(vjust = 0))
ggsave("figures_tables/missingdata_info.png", fig2a, height = 5, width =8)
```

Missingness only seen in outcome data (SAE) and more so for 2nd and third timepoints

Missingness by gender
```{r, fig.height = 3, fig.width=6, warning=FALSE}
tx_gender_missing <- combined_dataset |> 
  group_by(sex, time) |> 
  mutate(total = length(unique(id))) |> 
  ungroup() |> 
  drop_na() |> 
  group_by(sex, time) |> 
  summarize (missing_pct = (total[1] - length(unique(id)))*100/total[1])

missing_by_gender <- ggplot(aes(x = time, y = missing_pct, group = sex, color = sex), 
       data = tx_gender_missing) +
  geom_point() +
  scale_x_continuous(limits = c(1, 3))+
  geom_line() +
  labs(x = "Time after Vaccine (in months)", y = "Missing Data %", color = "Sex") +
  theme_classic()

ggsave("figures_tables/missingness_by_gender.pdf", missing_by_gender, height = 3, width =5)
#system("pdfcrop figures_tables/missingness_by_gender.pdf figures_tables/missingness_by_gender.pdf")
missing_by_gender
```

### Characterise missingness pattern
Number of drop outs 
```{r, warning=FALSE}
missingpattern <- combined_dataset_wide|> 
  mutate(pattern = case_when(is.na(month_3) == FALSE ~ "Complete", 
                             is.na(month_3) == TRUE & is.na(month_2) == TRUE &
                               is.na(month_1) == TRUE ~ "Assessment Dropout", 
                             is.na(month_3) == TRUE & is.na(month_2) == TRUE &
                               is.na(month_1) == FALSE ~ "Early Dropout", 
                             is.na(month_3) == TRUE ~ "Late Dropout"),  
         pattern = fct_relevel(pattern, "Assessment Dropout", "Early Dropout", 
                               "Late Dropout"))
missingpattern1 <- missingpattern |> 
  dplyr::select(c(sex, pattern)) |> 
  tbl_summary(
    by = pattern,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",    # drop min/max to save space
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      sex ~ "Sex")
  ) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels()

missingpattern1 |> 
  as_gt() |> 
  gt::gtsave(filename = "figures_tables/missingpattern1.pdf")

missingpattern1 |> 
  as_kable()
```

```{r, warning=FALSE}
missingpatternv2 <- missingpattern |> 
  dplyr::select(c(group, pattern)) |> 
  tbl_summary(
    by = group,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",    # drop min/max to save space
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      pattern ~ "Missingness Pattern")
  ) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels()

missingpatternv2 |> 
  as_gt() |> 
  gt::gtsave(filename = "figures_tables/missingpatternv2.pdf")

missingpatternv2 |> 
  as_kable()
```


Only 3.7% missingness, ignoring for now

EDA for SAE
```{r, fig.height = 4, fig.width=9, warning=FALSE}
sae_pct <- combined_dataset |> 
  drop_na() |> 
  group_by(group, time, site, sex) |> 
  mutate(total = n()) |> 
  ungroup() |> 
  dplyr::filter(sae == "Yes") |>  
  group_by(group, time, site, sex) |> 
  summarise (sae_pct = n()*100/total[1]) |> 
  mutate(group_time = paste(time, group, sep = "_")) 

sae_pct_by_gender_persite <- ggplot(aes(x = group_time, y = sae_pct, color = group), 
       data = sae_pct) +
  geom_boxplot() +
  geom_jitter(alpha = 0.4) +
  stat_compare_means(comparisons = list(c("1_Control", "1_Vaccine"), 
                                        c("2_Control", "2_Vaccine"), 
                                        c("3_Control", "3_Vaccine")), 
                     label = "p.adj.format",
                     p.adjust.method = "bonferroni") +
  labs(x = "Time after Vaccine (in months) + Tx Group", y = "Severe Adverse Event% (Per Site)", color = "Group") +
  facet_wrap(~sex) +
  theme_classic()

ggsave("figures_tables/sae_pct_by_gender_persite.pdf", sae_pct_by_gender_persite, height = 5, width =11)
#system("pdfcrop figures_tables/sae_pct_by_gender_persite.pdf figures_tables/sae_pct_by_gender_persite.pdf")
sae_pct_by_gender_persite
```

```{r, fig.height = 3, fig.width=5, warning=FALSE}
sae_pct <- combined_dataset |> 
  drop_na() |> 
  group_by(group, time) |> 
  mutate(total = n()) |> 
  ungroup() |> 
  dplyr::filter(sae == "Yes") |>  
  group_by(group, time) |> 
  mutate (sae_pct = n()*100/total[1])

sae_pct_by_gender <- ggplot(aes(x = time, y = sae_pct, group = group, color = group), 
       data = sae_pct) +
  geom_point() +
  scale_x_continuous(limits = c(1, 3), breaks = c(1, 2, 3))+
  geom_line() +
  labs(x = "Time after Vaccine (in months)", y = "Severe Adverse Event %", color = "Group") +
  theme_classic()

ggsave("figures_tables/sae_pct_by_gender.pdf", sae_pct_by_gender, height = 3, width =5)
#system("pdfcrop figures_tables/sae_pct_by_gender.pdf figures_tables/sae_pct_by_gender.pdf")
sae_pct_by_gender
```

```{r, fig.height = 3, fig.width=7, warning=FALSE}
sae_pct_by_group <- combined_dataset |> 
  drop_na() |> 
  group_by(sex, time, group) |> 
  mutate(total = n()) |> 
  ungroup() |> 
  dplyr::filter(sae == "Yes") |>  
  group_by(sex, time, group) |> 
  mutate (sae_pct = n()*100/total[1])

sae_pct_by_gender_group <- ggplot(aes(x = time, y = sae_pct, group = group, color = group), 
       data = sae_pct_by_group) +
  geom_point() +
  scale_x_continuous(limits = c(1, 3), breaks = c(1, 2, 3))+
  geom_line() +
  facet_wrap(~sex) +
  labs(x = "Time after Vaccine (in months)", y = "Severe Adverse Event %", color = "Group") +
  theme_classic()

ggsave("figures_tables/sae_pct_by_gender_group.pdf", sae_pct_by_gender_group, height = 3, width =7)
#system("pdfcrop figures_tables/sae_pct_by_gender_group.pdf figures_tables/sae_pct_by_gender_group.pdf")
sae_pct_by_gender_group
```


Modeling: GLMM to model for odds of SAE at each timepoint
```{r, warning=FALSE}
combined_dataset <- combined_dataset |> 
  mutate(time = as.factor(time))
glmm1 <- glmer(sae ~ group*time + group + sex + age + (1 | id), 
               family = binomial, data = combined_dataset, nAGQ = 0)
glmm2 <- glmer(sae ~ group + time + group + sex + age + (1 | id), 
               family = binomial, data = combined_dataset, nAGQ = 0)
anova(glmm1, glmm2)
```
Interaction term not needed

```{r}
combined_dataset <- combined_dataset |> 
  mutate(time = as.factor(time))
glmm3 <- glmer(sae ~ group*time + group + sex + age + (1 |site / id), 
               family = binomial, data = combined_dataset, nAGQ = 0)
anova(glmm1, glmm3)
```
read on correlation structure, addin two separate random intercept etc. 

